\chapter{The End of an Era, or is it?}

\section{Introduction}
\label{ch34-intro}%\hyperlabel{ch34-intro}%

In the last issue, we ended up with a
 LibGen\program{LibGen}\footnote{Well, yes we did, in the first version of the book. Since that was published, I've decided to rewrite the whole set of chapters on LibGen\program{LibGen} as a separate book, so watch this space....} application that was getting somewhere.
    But it's not finished yet. This issue might well be the last
 \emph{paper} copy of QL Today that you receive, but I have
    no current plans to stop development of this utility, nor to stop writing
    down stuff as I go along! I need to take a slight diversion into creating
    dynamic application sub-{}window menus before I can finish the utility
    properly. Unfortunately, this issue will not be continuing the program's
    development as I am in the middle of a huge amount of work in my current
    contract, and by the time you read this, I'll hopefully be in a new
    one.


\section{So What Now?}

Well, I have a half finished application and a few more articles on
    the Pointer Environment up my sleeve. Time permitting of course. As there
    will unlikely be a future paper version of QL Today, and I have no idea
    what the future of a replacement might be, I am setting up a mailing list
    on my web site so that anyone who wishes to take advantage of the
    remainder of the series, plus any other work I can think of and have time,
    to write, can.

You will be required to register on the list with a valid email
    address and I will also need your name too. My blog gets numerous
    registrations on a daily basis and most of them are from spam bots hoping
    to get free spam comments posted on my blog -{} they don't! Anyone signing
    up without a valid name and email gets deleted as part of my regular
    housekeeping. The mailing list will not allow you to register without a
    name and email address. Please supply your valid name. No nicknames
    please. That saves me some work clearing out the spam bots as well.

You will not get spammed by me when you register. Traffic will be
    light I imagine. Whenever I have an article ready, I'll send an email and
    supply a link where you can obtain the latest article. I'm looking at
    mailing list software that allows me to add attachments to the emails sent
    out, but so far, these seem few and far between -{} at least amongst the
    ones I'm allowed to use by my hosting company that is.

I know Dave Park mentioned that he would be setting up a Joomla
    system to replace the printed QL Today, but I haven't heard much for a
    while, so I'm not sure of progress on that matter.\footnote{As far as I am aware, this unfortunately, never happened. Dave's other commitments prevented him from getting this off the ground.}

Anyway, check the web page at \url{http://qdosmsq.dunbar-it.co.uk} to see
    if the details of the mailing list have been added, and if so, join up to
    keep reading the rest of the series. At least with my own mailing list,
    I'll have a half decent idea of how many readers I actually have!\footnote{At the last count, I have 54 readers!}
    ;-{})

\section{The End}
\label{ch34-the-end}%\hyperlabel{ch34-the-end}%

So, that's it. I've been writing these articles since the very first
    volume of QL Today, 17 years ago! It's been a long hard slog at times, and
    I haven't regretted a minute of it. I'd like to thank my faithful
    reader(s), George Gwilt who has far better coding skills than I have, and
    who kept a watchful eye on everything I wrote, offering corrections, bug
    fixes and observations on just about every article. Thanks George.

Hugh Rooms has commented on my articles as well as offering
    solutions too. And for that I'm grateful.

To all of you who read my articles and never once gave me any
    feedback, I thank you too. Without you, I wouldn't have as many readers as
    I have -{} but honestly, if you ever get involved in a series like that
    again, please give the author some feedback -{} even just a quick email to
    say "hello" or similar. Writing in isolation, for free, is fine, but it's
    far better to know that your efforts are being read by the "masses".
 

I wish everyone involved in QL Today, best wishes for the future.
\\
\\
\\
    Cheers, Norm.
